---
title: "R Notebook"
output: html_notebook
---

```{r}
### ENTER ROOT DIR PATH BELOW, or keep the renv/project rootdir
rootdir <- getwd()
Robjectsdir <- paste0(rootdir, "/Robjects")
Graphsdir <- paste0(rootdir, "/graphs")
BVdir <- paste0(Graphsdir, "/BioVar")


### ENTER PATH WITH KALLISTO FOLDERS BELOW
datapath <- paste0(rootdir, "/input")

# sample names for all samples
sample_list <- dir(file.path(datapath))
# paths to each kallisto folder for each processed sample
kal_dirs <- file.path(datapath, sample_list)

#setting root directory
knitr::opts_knit$set(root.dir = rootdir)
# options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = F, dplyr.summarise.inform = F) # avoid truncated output in R console and scientific notation
# Set seed to ensure reproducibility in any function sthat require randomness
set.seed(123456)
suppressMessages({
  library(edgeR)    # for DGE analysis
  library(pheatmap) # for heatmap graphs
  library(dplyr)    # for useful modification functions of dataframes, lists, etc
  library(data.table)      # to reorder the t2ged matrix
  library(clusterProfiler) # ORA/GSEA
  library(org.Dm.eg.db)    # Dmel db for GO terms/Kegg pathways
  library(DOSE)            # for some graphing functions (facet_grid)
  library(enrichplot)      # Graphing functions (pairwise_termsim)
  library(ggplot2)  # graphing module
  library(pathview) # to produce pathview plots of KEGG pathways with enriched genes/transcripts 
  library(cowplot)
  #library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
  library(RColorBrewer) # for a colourful plot
  library(ggrepel) # for nice annotations
  library(cluster)         # for PAM clustering function
  library(factoextra)      # for the various graphing methods in finding optimal k number
  library(openxlsx) # to save enrichment analysis results
  library(forcats) # to reverse order of factor for avglogCPM function
  library(tximport)
})
```

```{r loading relevant values}
if (!exists("y")){
  y <- readRDS(file = paste0(Robjectsdir, "/y_gene.rds"))
}
if (!exists("fit")){
  fit <- readRDS(file = paste0(Robjectsdir, "/fit.rds"))
}
if (!exists("s2c")){
  s2c <- readRDS(file = paste0(Robjectsdir, "/s2c.rds"))
  design <- s2c$design
}
if (!exists("t2g")){
  t2g <- readRDS(file = paste0(Robjectsdir, "/t2g_dmel.rds"))
}
if (!exists("t2g")){
  t2g <- readRDS(file = paste0(Robjectsdir, "/t2g_dmel.rds"))
}
if (!exists("logCPM")){
  logCPM <- readRDS(file = paste0(Robjectsdir, "/logCPM.rds"))
}




```



### Creating Contrast matrix containing all required contrasts to do DGE on

```{r Building contrasts matrix}
#####
#> To change:
#> - modify the contrasts matrix with all contrasts to potentially explore.
#> This section is very important as you are building the totality of contrasts to explore differential expression in. The contrasts matrix is later used for differential expression analysis, enrichment analysis, and to loop through other functions, as well as to label filenames and graphs with the contrast their data belongs to. 
#>
#####
#Building into one matrix for ease of use
contrasts <- makeContrasts(
  # Intra-host infection and injection contrasts
  WT_Inf_6h = W1118.U112.6h-W1118.PBS.6h,
  WT_Inf_48h = W1118.U112.48h-W1118.PBS.48h,
  WT_PBS_6h = W1118.PBS.6h-W1118.Unin.6h,
  WT_PBS_48h = W1118.PBS.48h-W1118.Unin.48h,
  IMD_Inf_6h = IMD.U112.6h-IMD.PBS.6h,
  IMD_Inf_48h = IMD.U112.48h-IMD.PBS.48h,
  IMD_PBS_6h = IMD.PBS.6h-IMD.Unin.6h,
  IMD_PBS_48h = IMD.PBS.48h-IMD.Unin.48h,
  # intra-host inf vs uninf
  WT_Inf_6h_nopbs = W1118.U112.6h-W1118.Unin.6h,
  WT_Inf_48h_nopbs = W1118.U112.48h-W1118.Unin.48h,
  IMD_Inf_6h_nopbs = IMD.U112.6h-IMD.Unin.6h,
  IMD_Inf_48h_nopbs = IMD.U112.48h-IMD.Unin.48h,
  
  # Inter-host infection and injection contrasts
  Inf_IMDvsWT_48h = (IMD.U112.48h-IMD.PBS.48h)-(W1118.U112.48h-W1118.PBS.48h),
  Inf_IMDvsWT_6h = (IMD.U112.6h-IMD.PBS.6h)-(W1118.U112.6h-W1118.PBS.6h),
  PBS_IMDvsWT_48h = (IMD.PBS.48h-IMD.Unin.48h)-(W1118.PBS.48h-W1118.Unin.48h),
  PBS_IMDvsWT_6h = (IMD.PBS.6h-IMD.Unin.6h)-(W1118.PBS.6h-W1118.Unin.6h),
  # time-independent intra-host contrasts
  # IMD_inf =
  # uninfected contrasts
  Uninf_IMDvsWT_6h = IMD.Unin.6h-W1118.Unin.6h,
  Uninf_IMDvsWT_48h = IMD.Unin.48h-W1118.Unin.48h,
  IMD_Uninf = IMD.Unin.48h-IMD.Unin.6h,
  WT_Uninf = W1118.Unin.48h-W1118.Unin.6h,
  # Time contrasts. Negative LFC means expression goes down over time, positive means it goes up
  WT_Inf = W1118.U112.48h-W1118.U112.6h,
  WT_PBS = W1118.PBS.48h-W1118.PBS.6h,
  IMD_Inf = IMD.U112.48h-IMD.U112.6h,
  IMD_PBS = IMD.PBS.48h-IMD.PBS.6h,
levels=s2c$group)

saveRDS(contrasts, file = paste0(Robjectsdir, "/contrasts.rds"))
```

### Running DGE analysis, producing relevant graphs, and saving data to Robject file
Differential Expression testing

Using glmTreat function to perform Differential expression testing with a log-fold change cutoff of 1.5. With input being a DGEGLM object from glmQLFit(), the testing method used is the same as that used in glmQLFTest, that is a quasi-likelihood F-test, but with the null hypothesis tested as the absolute value of the logFC of each transcript being equal to or below 1.5.

```{r DE testing}
#####
#> To change:
#> - After getting the final number of DE transcripts for each contrast, determine if a significance cut-off is desired. 
#> Remember that any cut-off will bias the resulting significantly DE transcripts for transcripts with potentially low overall expression but relatively high differences in expression in the contrast. Transcripts with higher average expression will need larger increase/decrease in absolute expression values in the contrast to pass the filter, which can hide DE transcripts of genes that are usually highly expressed. But also, getting very large number of DE transcripts (in proportion to total number of transcripts in genome/counts) does inform the need for a cut-off.
#>
#####

# takes as input the DGEGLM object fit previously made when calling glmQLFit. Default FDR is at 5%. Using glmTreat to allow for log-fold-cutoff to focus on more biologically relevant transcripts with higher fold-change in the tested contrast
Treat_res <- list()
for (i in 1:dim(contrasts)[2]){
  Treat_res[[colnames(contrasts)[i]]] <- glmTreat(fit, contrast = contrasts[,i], lfc = log2(1.2))
}

# same method without log-fold cutoff:
glmQL_res <- list()
for (i in 1:dim(contrasts)[2]){
  glmQL_res[[colnames(contrasts)[i]]] <- glmQLFTest(fit, contrast = contrasts[,i])
}
# tximport method labels gene ID column names differently to catch kallisto. this changes them to be identical for ease of use when coding
if (!is.null(glmQL_res[[1]]$genes$TranscriptID)){
  for (i in 1:length(glmQL_res)){
    colnames(glmQL_res[[i]]$genes)[colnames(glmQL_res[[i]]$genes) == 'TranscriptID'] <- 'target_id'
    colnames(glmQL_res[[i]]$genes)[colnames(glmQL_res[[i]]$genes) == 'flybaseID'] <- 'ens_gene'
    colnames(glmQL_res[[i]]$genes)[colnames(glmQL_res[[i]]$genes) == 'ENTREZID'] <- 'entrez_gene'
    colnames(glmQL_decided[[i]]$table)[colnames(glmQL_decided[[i]]$table) == 'TranscriptID'] <- 'target_id'
    colnames(glmQL_decided[[i]]$table)[colnames(glmQL_decided[[i]]$table) == 'flybaseID'] <- 'ens_gene'
    colnames(glmQL_decided[[i]]$table)[colnames(glmQL_decided[[i]]$table) == 'ENTREZID'] <- 'entrez_gene'
  }
}

saveRDS(glmQL_res, file = paste0(Robjectsdir, "/glmQL_res_txi.rds"))
saveRDS(Treat_res, file = paste0(Robjectsdir, "/Treat_res_txi.rds"))

```
From above DE testing, number of significantly up/downregulated transcripts, and non-significant transcripts. Looking at 48h inter-host infection contrast here:

```{r Outputting number of DE transcripts, include=FALSE}
glmQL_decided <- list()
glmQL_decided_numbs <- list()

for (i in 1:length(glmQL_res)){
  # multiple testing verification to classify trasncripts as significantly upregulated/downregulated or nonsignificant.
  glmQL_decided_numbs[[names(glmQL_res)[i]]] <- decideTests(glmQL_res[[i]])
  glmQL_decided[[names(glmQL_res)[i]]] <-  topTags(glmQL_res[[i]], n= nrow(glmQL_res[[i]]))
}

Treat_decided <- list()
Treat_decided_numbs <- list()

for (i in 1:length(Treat_res)){
  Treat_decided_numbs[[names(Treat_res)[i]]] <- decideTests(Treat_res[[i]])
  Treat_decided[[names(Treat_res)[i]]] <- topTags(Treat_res[[i]], n= nrow(Treat_res[[i]]))
}
saveRDS(glmQL_decided, file = paste0(Robjectsdir, "/glmQL_decided_txi.rds"))
saveRDS(Treat_decided, file = paste0(Robjectsdir, "/Treat_decided_txi.rds"))


```

```{r summaries}
DGEs <- list()
for (i in 1:length(glmQL_decided)){
  DGEs[[names(glmQL_res)[i]]] <- summary(glmQL_decided_numbs[[names(glmQL_res)[i]]])
}
DGEs_Treat <- list()
for (i in 1:length(Treat_decided)){
  DGEs_Treat[[names(Treat_decided)[i]]] <- summary(Treat_decided_numbs[[names(Treat_decided)[i]]])
}
DGEs
DGEs_Treat
```



